Vagueness, Noise, and Communication

### 2014-10-02 paper/01_intro.tex
introduction and background; sec com

Many of our concepts and words are vague. A vague category knows clear
cases that fall under it, clear cases that do not, and also so-called
borderline cases in-between that neither clearly apply, nor clearly
not apply, but might themselves not be equal in terms of how well they
represent the category in question. Vagueness does not seem to
dramatically affect the success of everyday communication, but it is
troublesome for some of our best theories of language. Especially the
logico-semantic tradition of Frege, Russell and the young Wittgenstein
is challenged by vagueness and the paradoxes that it gives rise to. As
can be expected, many proponents of this tradition have risen to the
challenge \citep{Williamson1994:Vagueness}.

There are other intriguing aspects about vagueness. Linguists, for
example, wonder about the reasons why some expressions are less vague
than others \citep[e.g.][]{Kennedy2007:Vagueness-and-G}. But the
puzzle that we are concerned with here is how vagueness could arise
and be maintained in the first place. This may not seem a particularly
deep issue at first glance, but on closer inspection it is a serious
worry to functionalist accounts that maintain that our concepts and
language use evolved in order to be efficient. Since the existence of
unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward
by \citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
exist despite its obvious and demonstrable functional deficiency under
evolutionary pressure to be optimally expressive.

This problem, call it Lipman's problem, has a conceptual and a
technical side to it. Conceptually, it asks for reasons and general
mechanisms by which we could plausibly conceive of vagueness as
persisting the pressure of evolutionary optimization towards
precision. Such reasons and general mechanisms are, arguably, not hard
to imagine. We will address some presently. But each putative
explanation, no matter how plausible intuitively, must also be checked
for internal consistency and its potential to account for the fact
that we see vagueness as a part of a by-and-large efficient system of
categorization and communication. In other words, not all first-shot
rebuttals of Lipman's problem will do. Denying that there is any
functional pressure on efficient communication, for instance, is
dubious because it leaves the relative efficiency of our communication
with vague words entirely unexplained. In sum, the technical answer to
Lipman's problem is to give, ideally, a model of evolution of
categorization or language use that shows how competing forces that
realize both by-and-large efficient categories and vagueness.

c

@@

\begin{itemize}
\item introduce sim-max games
\item problem of vagueness \citep{Lipman2009:Why-is-Language}
\item \dots
\end{itemize}

\subsection{Situating our approach}

A number of authors have recently tried to explain why vagueness
evolved, based on considerations why a vague language might be useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,Jaegherde-JaegherRooijvan-Rooij2010:Strategic-Vague,BlumeBoard2013:Intentional-Vag}. In
contrast to these, others have argued that vagueness if a natural
byproduct of limitations in information processing
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This
paper makes a contribution to the latter line of thought.

The view that vagueness is a natural concomitant of cognitive
limitations of language users has been formalized in a number of
ways. As in other applications of evolutionary game theory, we can
distinguish micro- and macro-level approaches. Micro-level approaches
look at adaptive behavior of individual agents. Usually, changes in
the behavioral dispositions of agents occur after every single
interaction. In contrast, macro-level approaches outline more
abstract, aggregate population-based dynamics. Usually, a macro-level
dynamic captures changes of frequencies of behavioral types in the
population over time. Previous accounts have suggested both micro- and
macro-level dynamics to show how vague signal use can arise in sim-max
games.

@@

To understand the workings of the \rdd, we study numerical simulations
\dots

The paper is structured as follows:
\begin{enumerate}
\item replicator diffusion dynamics
\item relation to replicator mutator dynamics
\item simulations of \rdd
\item comparison with other approaches
\item conclusions
\end{enumerate}

>>>>

Many of our concepts and words are vague. A vague category knows clear
cases that fall under it, clear cases that do not, and also so-called
borderline cases in-between that neither clearly apply, nor clearly
not apply, but might themselves not be equal in terms of how well they
represent the category in question. Vagueness does not seem to
dramatically affect the success of everyday communication, but it is
troublesome for some of our best theories of language. Especially the
logico-semantic tradition of Frege, Russell and the young Wittgenstein
is challenged by vagueness and the paradoxes that it gives rise to. As
can be expected, many proponents of this tradition have risen to the
challenge
\citep[e.g.][]{Williamson1994:Vagueness,CobrerosEgre2012:Tolerant-Classi}.\todo{more
  references}

Yet there are other intriguing aspects about vagueness. The
puzzle that we are concerned with here is how vagueness could arise
and be maintained in the first place. This may not seem a particularly
deep issue at first glance, but on closer inspection it is a serious
worry to functionalist accounts that maintain that our concepts and
language use evolved in order to be efficient. Since the existence of
unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward
by \citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
exist despite its obvious and demonstrable functional deficiency under
evolutionary pressure to be optimally expressive.

This problem, call it Lipman's problem, has a conceptual and a
technical side to it. Conceptually, it asks for reasons and general
mechanisms by which we could plausibly conceive of vagueness as
resisting the pressure of evolutionary selection for precision. Such
reasons and general mechanisms are, arguably, not hard to imagine. We
will address some presently. But each putative explanation, no matter
how plausible intuitively, must also be checked for internal
consistency and its potential to account for the fact that we see
vagueness as a part of a by-and-large efficient system of
categorization and communication. In other words, not all first-shot
rebuttals of Lipman's problem will do. Denying that there is any
functional pressure on efficient communication, for instance, is a
non-starter, because it leaves the relative efficiency of our
communication with vague words entirely unexplained. In sum, the
technical answer to Lipman's problem is to give, ideally, a model of
evolution of categorization or language use that shows how competing
forces that realize both by-and-large efficient categories and
vagueness.

A number of authors have recently tried to explain why vagueness
evolved, based on considerations why a vague language might be useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,Jaegherde-JaegherRooijvan-Rooij2010:Strategic-Vague,BlumeBoard2013:Intentional-Vag}. In
contrast to these, others have argued that vagueness is a natural
byproduct of limitations in information processing
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o,QingFranke2014:Gradable-Adject}. This
paper makes a contribution to the latter line of thought.  Concretely,
we introduce a novel variant of the replicator mutator dynamic that
integrates stochastic noise on the differential confusability of
similar stimuli. Our main contribution, therefore, is a technical
answer to Lipman's problem in the sense introduced above. The dynamic
proposed here nicely improves on and complements previous like-minded
accounts. We show that that stochastic noise can not only lead to
vague, yet communicative efficient signal use, but can also unify
evolutionary outcomes and help avoid inefficient categorization.

The next section introduces the background against which the work
presented here can be appreciated. Section~\ref{sec:repl-diff-dynam},
then, introduces the replicator diffusion dynamic, and elaborates its
relation with the replicator mutator
dynamic. Section~\ref{sec:exploring-rdd} explores the replicator
diffusion dynamic on the relevant class of generalized signaling games
introduced in
Section~\ref{sec:background}. Section~\ref{sec:discussion} reflects on
the results and compares what has been achieved to related accounts in
more detail.

\section{Background}
\label{sec:background}

The view that vagueness is a natural concomitant of cognitive
limitations of language users has been formalized in a number of ways,
using evolutionary game theory and certain generalizations of
signaling games, so called similarity-maximizing games or sim-max
games, for short
\citep{FrankeJager2010:Vagueness-Signa,OConnor2013:Evolving-Percep,OConnor2013:The-Evolution-o}. Our
contribution is best seen in relation to these accounts, as it also
relies on sim-max games. Let's introduce these first.

Signaling games, as introduced by \citet{Lewis_1969:Convention}, have
a sender and a receiver. The sender knows the true state of the world,
but the receiver does not. The sender can select a signal, or message,
to reveal to the receiver, who then chooses an act. In Lewis' games,
states are maximally distinct from each other and the receiver's acts
are related to them one-to-one. If the receiver chooses the act that
corresponds to the actual state, the play is a success, otherwise a
failure. Similarity-maximizing games are generalizations of Lewis'
games in that they allow different states to be more or less similar
to one another, and, roughly put, to make success of communication a
function of that similarity. Signaling games with utility-relevant
similarities in the state space are fairly standard in economics
\citep[e.g.][]{Spence1973:Job-market-sign,CrawfordSobel1982:Strategic-Infor},
but have received particular attention in a more philosophical and
linguistic context by the more recent work of
\citet{Jager2007:The-Evolution-o,JagerRooijvan-Rooij2007:Language-Struct,JagerMetzger2011:Voronoi-Languag}.

More concretely, a sim-max game, in the relevant sense here, consists
of a set of states $\States$, a set of messages $\Messgs$, a
similarity metric on states: $\similarity \mycolon \States \times
\States \rightarrow \mathds{R}$, and a utility function $\Utils
\mycolon \States \times \States \rightarrow \mathds{R}$. We identify
the receiver's acts with the states of the world, so that the game is
one of guessing the actual state, so to speak. We also assume that
sender and receivers interests are alike, so we only have one utility
function. We do not consider message costs, so utilities only depend
on the actual state, and the receiver's response. The similarity function
should satisfies the usual requirements for a metric:
\begin{align*}
  & \similarity(\mystate{1}, \mystate{2}) \ge 0 &&
  \similarity(\mystate{1}, \mystate{2}) = 0 \Leftrightarrow
  \mystate{1} = \mystate{2} \\
  & \similarity(\mystate{1}, \mystate{2}) = \similarity(\mystate{2},
  \mystate{1}) && \similarity(\mystate{1}, \mystate{2}) \le
  \similarity(\mystate{1}, \mystate{3}) + \similarity(\mystate{3}, \mystate{2})\,.
\end{align*}
The utility function should be a monotonically decreasing function of
similarity:
\begin{align*}
  \similarity(\mystate{1},\mystate{2}) \ge
  \similarity(\mystate{1},\mystate{3}) \ \ \Rightarrow \ \ 
  \Utils(\mystate{1},\mystate{2}) \ge
  \Utils(\mystate{1},\mystate{3})\,.
\end{align*}
To keep matters simple, we only look at cases where $\States$ contains
finitely many points from the unit interval, with similarity given by
the Euclidean distance in this one-dimensional space.\todo{introduce
  utilities here?}

\begin{itemize}
\item ESSs of sim-max games
\item Voronoi tessallations
\item conceptual spaces
\item 
\end{itemize}

@@

As in other applications of
evolutionary game theory, we can distinguish micro- and macro-level
approaches. Micro-level approaches look at adaptive behavior of
individual agents. Usually, changes in the behavioral dispositions of
agents occur after every single interaction. In contrast, macro-level
approaches outline more abstract, aggregate dynamics, happening in a
huge population of agents, or otherwise abstracting from seemingly
irrelevant detail. Usually, a macro-level dynamic captures changes of
frequencies of behavioral types in the population over time. Previous
accounts have suggested both micro- and macro-level dynamics to show
how vague signal use can arise in sim-max games.

@@

### 2014-10-02 paper/04_discussion.tex
introduction and background; sec com

Diffusion of strategies in the \rdd can be conceived of as a form of
generalization as well, and works in large part quite analogous to
stimulus generalization in \citeauthor{OConnor2013:The-Evolution-o}'s
approach. This holds for the effect of diffusion and generalization,
but not necessarily for the way that the effect is achieved. We also
saw that diffusion in \rdd leads to more regular languages and
speedier convergence. However, the \rdd is a more abstract framework
than generalized reinforcement learning. The latter is foremost
motivated as a learning dynamic that has two players adapt their
individual strategies after each concrete round of play. In contrast,
the \rdd describes a more abstract, average dynamical change in
behavioral dispositions. Although the dynamics of (some forms of)
reinforcement learning mirror those of the replicator dynamics (at
some stage in time) (\cite{Borgers1997,Hopkins2005,Beggs2005}), this
does not mean that \emph{generalized} reinforcement learning, in which
stimulus generalization is best motivated at the level of a single
agent's dispositional generalization after one round of play, is also
directly a plausible high-level description of, say, generalized
learning in a population of agents. Seen in this light, generalized
reinforcement learning and the \rdd nicely complement each other, as
similarly-minded accounts operating at different levels of
abstraction.

@@

>>>>

Diffusion of strategies in the \rdd can be conceived of as a form of
generalization as well, and works in large part quite analogous to
stimulus generalization in \citeauthor{OConnor2013:The-Evolution-o}'s
approach. This holds for the effect of diffusion and generalization,
but not necessarily for the way that the effect is achieved. We also
saw that diffusion in \rdd leads to more regular languages and
speedier convergence. However, the \rdd is a more abstract framework
than generalized reinforcement learning. The latter is foremost
motivated as a learning dynamic that has two players adapt their
individual strategies after each concrete round of play. In contrast,
the \rdd describes a more abstract, average dynamical change in
behavioral dispositions. Although the dynamics of (some forms of)
reinforcement learning mirror those of the replicator dynamics (at
some stage in time)
(\cite{BorgersSarin997:Learning-Throug,HopkinsPosch2005:Attainability-o,Beggs2005:On-the-Converge}),
this does not mean that \emph{generalized} reinforcement learning, in
which stimulus generalization is best motivated at the level of a
single agent's dispositional generalization after one round of play,
is also directly a plausible high-level description of, say,
generalized learning in a population of agents. Seen in this light,
generalized reinforcement learning and the \rdd nicely complement each
other, as similarly-minded accounts operating at different levels of
abstraction.

@@

\paragraph{The conceptual spaces approach.}

\begin{itemize}
\item can only ``construct'' (fuzzy) tessellation's given the
  prototypes (or prototype regions); but cannot explain how the
  prototypes come about
\end{itemize}

### 2014-10-02 paper/paper.tex
data exploration MF
introduction; iterated diffusion
improved structure Sec 02; data for convex vs. non-convex strats for small state spaces
restructured and expanded Section 2
iterated diffusion; plots
more on diffusion; Nosofsky similarity plots
diffusion section ready
more data exploration with varying tolerance and impairment
simulation results write up
results report
introduction and background; sec com

\usepackage[]{../helpers/mypackages}
%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}
\bibliography{../helpers/MyRefGlobal}
\bibliography{paper}
\usepackage{../helpers/myenvironments}
\usepackage{../helpers/mycommands}
\usepackage{todonotes}
\usepackage{subcaption}

@@

\title{Vagueness, Noise, and Communication}
\author{Jos\'e Pedro Correia and Michael Franke}
\date{}

@@

\begin{abstract}
Signaling games have been garnering a lot of attention as models for studying natural language meaning.
One important aspect that is nevertheless typically missing from most models in the literature is the possibility for vagueness as a feature of successful languages.
Recently, some models have been proposed to explicitly address this limitation.
In this paper we explore one such model and establish a link between the dynamics in the model and the replicator-mutator equations.
Furthermore, we study the long term behavior of the model using computer simulations.
The results indicate that vague languages, despite having lower expected utility and informativity, are faster to converge to an equilibrium than non-vague ones.
We argue that this could be an importante advantage that contributes to the pervasiveness of vagueness in natural language.
\end{abstract}

>>>>

\usepackage[]{../helpers/mypackages}
%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}
\bibliography{MyRefGlobal}
% \bibliography{paper}
\usepackage{../helpers/myenvironments}
\usepackage{../helpers/mycommands}
\usepackage{todonotes}
\usepackage{subcaption}

@@

\newcommand{\rd}{\acro{rd}} % replicator dynamic
\newcommand{\rmd}{\acro{rmd}} % replicator mutator dynamic
\newcommand{\rdd}{\acro{rdd}} % replicator diffusion dynamic
\newcommand{\RD}{\ensuremath{\mathrm{RD}}} % replicator dynamic
\newcommand{\RDD}{\ensuremath{\mathrm{RDD}}} % replicator diffusion dynamic
\newcommand{\RMD}{\ensuremath{\mathrm{RMD}}} % replicator mutator

\newcommand{\Diff}{\ensuremath{\mathrm{D}}} % Difusion 
\newcommand{\Mutate}{\ensuremath{\mathrm{M}}} % Mutation

\newcommand{\impairment}{\ensuremath{\alpha_i}} % impairment
\newcommand{\toler}{\ensuremath{\alpha_t}} % tolerance
\newcommand{\ns}{\ensuremath{n_s}} % number of states

\newcommand{\similarity}{\ensuremath{\mathrm{Sim}}} % similarity function

\title{Vagueness, Noise, and Communication}
% \author{Jos\'e Pedro Correia and Michael Franke}
\date{}

@@

\begin{abstract}
  Signaling games have attracted a lot of attention as models for
  studying the evolution of meaning. One important aspect that is
  nevertheless typically missing from most models in the literature is
  the possibility for vagueness as a feature of successful signaling.
  Complementing recent models that have been proposed to explicitly
  address this limitation, we introduce the replicator diffusion
  dynamic, provably a special case of the replicator mutator dynamic,
  which implements a versatile and natural integration of stochastic
  noise in the form of probabilistic confusion of similar stimuli, on
  the one hand, and evolutionary pressure on optimal signal use, on
  the other. Applying the replicator diffusion dynamic to a
  generalization of David Lewis' signaling games, so-called
  similarity-maximizing games, we show that stochastic noise can not
  only lead to vague, yet communicative efficient signal use, but can
  also unify evolutionary outcomes and help avoid inefficient
  categorization.
\end{abstract}
